---
title: " rarefaction toy-example "
author: "Frédéric Mahé"
date: '`r format(Sys.time(), "%d %B %Y")`'

output:
  rmarkdown::html_document:
    theme: lumen
    toc: yes
    toc_float: TRUE
    keep_md: yes
    # code_folding: hide
---

```{r setup, include=FALSE}
rm(list = ls()) # remove all objects before starting
knitr::opts_chunk$set(echo = TRUE)
```

***

#### load required packages

```{r packages, message=FALSE}
library(tidyverse)
library(vegan)
```


# preliminary checks

## variables and functions

```{r}
fake_abundances <- c(30, 20, 15, 10, 7, 5, 5, 3, 2, 1, 1, 1, 0)
min_number_of_reads <- 10
seed <- 123
n_subsamplings <- 1000
```

```{r}
rarefy_n_times <- function(transposed_table, sample_size, n) {
    set.seed(seed)

    transposed_table %>%
        vegan::rrarefy(x = .,
                       sample = sample_size) -> local_matrix

    for (i in 2:n) {
        local_matrix <-
            local_matrix +
            vegan::rrarefy(x = transposed_table,
                           sample = sample_size)
    }

    return(local_matrix / n)
}


diversity_index_n_times <- function(transposed_table, sample_size, n, diversity_index) {
    set.seed(seed)

    transposed_table %>%
        vegan::rrarefy(x = .,
                       sample = sample_size) %>%
        vegan::diversity(x = .,
                         index = diversity_index) -> index_value

    for (i in 2:n) {
        transposed_table %>%
            vegan::rrarefy(x = .,
                           sample = sample_size) %>%
            vegan::diversity(x = .,
                             index = diversity_index) -> local_index_value

        index_value <- index_value + local_index_value
    }

    return(index_value / n)
}


dissimilarity_n_times <- function(transposed_table, sample_size, n, diversity_index) {
    set.seed(seed)

    replicate(n = n,
              expr = transposed_table %>%
                  vegan::rrarefy(x = ., sample = sample_size) %>%
                  vegan::vegdist(x = ., index = diversity_index)) %>%
        sum() -> local_sum

    return(local_sum / n)
}
```


## data creation and preliminary checks

here we go!

```{r}
tibble(s1 = fake_abundances) %>%
    t() -> raw_table_transposed
```

There is one sample with `r fake_abundances %>% sum(.) %>%
prettyNum(., scientific=FALSE, big.mark=",")` reads.

As we rarefy to `r min_number_of_reads %>% prettyNum(.,
scientific=FALSE, big.mark=",")` we would expect our fake abundance values to be divided by `r (fake_abundances %>% sum(.) / min_number_of_reads) %>% prettyNum(., scientific=FALSE, big.mark=",")`:


```{r}
tibble(s1 = fake_abundances) %>%
    mutate(s1 = s1 / (sum(s1) / min_number_of_reads)) %>%
    pull() -> expected_abundance_values_after_rarefaction
```

Let's see how close we can get with an *averaged rarefaction*.
 

# rarefaction (random subsampling)

Randomly subsample the table, so all samples have the same number of
reads. Repeat the process `r n_subsamplings` times to make sure that
the final profile is as close as possible to the initial
distribution. Use a fix seed to make the process 100% repeatable.

## rarefy once

Keep `matrix1` as a reference:

```{r}
set.seed(seed)

raw_table_transposed %>%
    vegan::rrarefy(x = .,
                   sample = min_number_of_reads) -> matrix1

matrix1 -> first_rarefaction
```

How far are we from our expectations?

```{r}
first_rarefaction %>%
    t() %>%
    as_tibble() %>%
    pull()

expected_abundance_values_after_rarefaction %>%
    round(x = .)
```

Close but the fact that rarefaction is inherently a random process is
obvious. We might want to limit random dispersion by computing *n*
independent rarefaction, adding them and dividing by *n* to get
averaged rarefaction results. Doing so produces tables of decimal
values, rather than integral values. A potential issue is that
diversity toolboxes such as `vegan` may not accept and process decimal
values correctly.

We need to verify if diversity toolboxes such as vegan can compute
alpha- and beta-diversity stats using decimal decimal values, or if it
is necessary to round up the abundance values.

We already know that `vegan::estimateR()` requires integral values,
and that `vegan::specnumber()` overestimates the number of species
when given decimal values. So rounding is already necessary for some
stats.

A potential issue is that rounding slightly changes the actual number
of reads per sample.


## rarefy *n* times and compute average

```{r}
rarefy_n_times(transposed_table = raw_table_transposed,
               sample_size = min_number_of_reads,
               n = n_subsamplings) -> raw_table_transposed_rarefied
```

How far are we from our expectations?

```{r}
raw_table_transposed_rarefied %>%
    t() %>%
    as_tibble() %>%
    pull() -> averaged_rarefied_values

averaged_rarefied_values

averaged_rarefied_values %>%
    round(x = .)

expected_abundance_values_after_rarefaction %>%
    round(x = .)
```

The more repeats, the closer we are.

Notice that the sum of reads is correct before rounding up, but
smaller than expected after `round()`:

```{r}
averaged_rarefied_values %>%
    sum()

averaged_rarefied_values %>%
    round(x = .) %>%
    sum()
```

As stated previously, this is a potential issue for beta-diversity
analysis.


# compute alpha-diversity stats

## estimateR

```{r}
first_rarefaction %>%
    vegan::estimateR(x = .)

raw_table_transposed_rarefied %>%
    round(x = .) %>%
    vegan::estimateR(x = .)
```

Rounding is mandatory. Results between a single-rarefaction and an
averaged rarefaction are close.


```{r}
raw_table_transposed %>%
    vegan::estimateR(x = .)
```

Working directly with the original non-rarefied data gives different
results, as expected.


## specnumber

```{r}
first_rarefaction %>%
    vegan::specnumber(x = .)

raw_table_transposed_rarefied %>%
    vegan::specnumber(x = .)

raw_table_transposed_rarefied %>%
    round(x = .) %>%
    vegan::specnumber(x = .)
```

Rounding is not mandatory, but necessary.


## diversity

Here we introduce the `diversity_index_n_times()` function that can
rarefy a matrix and compute a diversity index *n* times independently,
and finally return an average value. This is our most precise
evaluation of the true index value.


### Shannon

```{r}
first_rarefaction %>%
    vegan::diversity(., index = "shannon")

raw_table_transposed_rarefied %>%
    vegan::diversity(., index = "shannon")

raw_table_transposed_rarefied %>%
    round(x = .) %>%
    vegan::diversity(., index = "shannon")

diversity_index_n_times(transposed_table = raw_table_transposed,
                        sample_size = min_number_of_reads,
                        n = n_subsamplings,
                        diversity_index = "shannon")

raw_table_transposed %>%
    vegan::diversity(x = ., index = "shannon")
```

Results with the averaged rounded table are very close to the results
computed independently *n* times and averaged (more expensive
computation).

```{r}
raw_table_transposed %>%
    vegan::diversity(x = ., index = "shannon")

raw_table_transposed_rarefied %>%
    vegan::diversity(., index = "shannon")
```

The original non-rarefied data and the rarefied averaged non-rounded
data are close. This is expected: without rounding the two
distributions are the same, just scaled differently (see below). It
also shows that the `vegan::diversity()` function accepts real
numbers.

```{r}
raw_table_transposed

raw_table_transposed_rarefied
```


### Inverse Simpson

```{r}
first_rarefaction %>%
    vegan::diversity(., index = "invsimpson")

raw_table_transposed_rarefied %>%
    vegan::diversity(., index = "invsimpson")

raw_table_transposed_rarefied %>%
    round(x = .) %>%
    vegan::diversity(., index = "invsimpson")

diversity_index_n_times(transposed_table = raw_table_transposed,
                        sample_size = min_number_of_reads,
                        n = n_subsamplings,
                        diversity_index = "invsimpson")

raw_table_transposed %>%
    vegan::diversity(x = ., index = "invsimpson")
```

Results with the averaged rounded table are very close to the results
computed independently *n* times and averaged (more expensive
computation).

```{r}
rm(raw_table_transposed)
```


## conclusion

Results obtained with the rarefied-averaged rounded table are valid,
at least for alpha-diversity.


# compute beta-diversity stats

Create a toy-example with two dissimilar samples:

```{r}
tibble(s1 = fake_abundances,
       s2 = rev(fake_abundances)) %>%
    t() -> raw_table_transposed
```

Bray-Curtis dissimilarity with and without rarefaction:

without rarefaction:

```{r}
raw_table_transposed %>%
    vegan::vegdist(x = ., method = "bray")
```

with *n* independent rarefactions and dissimilarity assessments:

```{r}
dissimilarity_n_times(transposed_table = raw_table_transposed,
                      sample_size = min_number_of_reads,
                      n = n_subsamplings,
                      diversity_index = "bray")
```

with *n* independent rarefactions and one dissimilarity assessment, without and with rounding:

```{r}
rarefy_n_times(transposed_table = raw_table_transposed,
               sample_size = min_number_of_reads,
               n = n_subsamplings) -> raw_table_transposed_rarefied
```

```{r}
raw_table_transposed_rarefied %>%
    vegan::vegdist(x = ., method = "bray")

raw_table_transposed_rarefied %>%
    round() %>%
    vegan::vegdist(x = ., method = "bray")
```

## conclusion

- as expected, rarefaction reduces precision on the measured
  dissimilarity (0.79 -> 0.8919),
- rarefying and averaging dissimilarities, without rounding, yields a
  dissimilarity value close to the one obtained on the original
  non-rarefied data (0.79 vs 0.7916),
- with rounding, rarefying and averaging dissimilarities yields a
  higher (maximal) dissimilarity value of 1.0. After rarefaction, our
  fake samples do not overlap, so a maximal dissimilarity is expected,
- when running more independent rarefactions and dissimilarity
  assessments, the observed average dissimilarity value seems to
  remain stable around 0.89

Now, what should we do? Rarefaction brings a lot of issues, but it is
necessary for now if we want to compare samples. So, it seems that the
ideal way is to repeat *n* independent rarefactions and diversity
assessments, at least for alpha statistics). This is computationally
costly. Our recently introduced strategy of averaging *n* independent
rarefactions saves some computation time, but introduces new
difficulties. However, rarefying and averaging dissimilarities without
rounding, yields a dissimilarity value close to the one obtained on
the original non-rarefied data.

In short: results obtained with the rarefied-averaged **not** rounded
table are valid, at least for the Bray-Curtis dissimilarity.


## long-term solution?

A potential solution could be to rarefy n times (or to divide
abundances by a factor), then to round-up, then to rarefy again only
once on the size of the smallest sample. That process would yield a
table of integers, very close to the optimal rarefaction, with the
same number of reads in all samples.


# not-so-good solutions

Decimal shifting (multiple by a factor): 

```{r}
(raw_table_transposed_rarefied * n_subsamplings) %>%
    vegan::estimateR(x = .)

(raw_table_transposed_rarefied * n_subsamplings) %>%
    vegan::specnumber(x = .)
```

Inflation of the number of species, not good.


***

```{r}
sessionInfo()
rm(list = ls())
```
